Dependent Object Types

### 2012-08-08 article.tex
spellcheck

This paper presents a proposal for a new type-theoretic foundation of
Scala and languages like it. The properties we are interested in
modelling are Scala's path-dependent types and abstract type members,
as well as its mixture of nominal and structural typing through the
use of refinement types. Compared to previous approaches~\cite{nuObj,FS},
we make no attempt to model inheritance or mixin composition. Indeed
we will argue that such concepts are better modelled in a different
setting.

The calculus does not precisely describe what's currently in Scala. It
is more normative than descriptive. The main point of deviation
concerns the difference between Scala's compound type formation using
{\bf with} and classical type intersection, as it is modelled in the
calculus. Scala, and the previous calculi attempting to model it
conflates the concepts of compound types (which inherit the members of
several parent types) and mixin composition (which build classes from
other classes and traits). At first glance, this offers an economy of
concepts. However, it is problematic because mixin composition and
intersection types have quite different properties. In the case of
several inherited members with the same name, mixin composition has to
pick one which overrides the others. It uses for that the concept of
linearization of a trait hierarchy. Typically, given two independent
traits $T_1$ and $T_2$ with a common method $m$, the mixin composition
\code@$T_1$ with $T_2$@ would pick the $m$ in $T_2$, whereas the member in
$T_1$ would be available via a super-call. All this makes sense from
an implementation standpoint. From a typing standpoint it is more
awkward, because it breaks commutativity and with it several
monotonicity properties.

@@

Sets of declarations form a lattice with the given meet $\wedge$ and
join $\vee$, the empty set of declatations as the top element, and the
bottom element $\seq{D_\Bot}$, Here $\seq{D_\Bot}$ is the set of
declarations that contains for every term label $l$ the declaration
$\ldecl l \Bot$, for every type label $L$ the declaration $\Ldecl L
\Top \Bot$ and for every method label $m$ the declaration $\mdecl m
\Top \Bot$.
}

@@

\subsubsection{Set of Terms}
$\rele k \Gamma s T$ defines the set of terms that appear to have type
$T$ when taking at most $k$ steps. $s$ must agree with a {\it prefix}
of $\Gamma$, so $\Gamma$ can additionaly contain variables not in
$s$. This is needed for checking methods in $\mathcal{V}$ above, and
for relating open terms. If $k > 0$, $\mathcal{E}$ extends $\Gamma$
and $s$ so that they agree. It then states that if it can reduce $t$
in the extended store to an irreducible term in $j < k$ steps, then
this term must be in a corresponding $\mathcal{V}$ set.
\begin{align*}
&\rele k \Gamma s T = \{ t \ |\\
&\gap \Gamma \ts T \wfe \andl\\
&\gap k = 0 \orl (\forall j \leq k,\\
&\gap\gap \forall (\Gamma'; s') \in \rels k \Gamma s ,\\
&\gap\gap \reductionl t {s'} {t'} {s''} j \andl\\
&\gap\gap \irred {t'} {s''} \impliesl\\
&\gap\gap \forall \Gamma'' \in \relg k {s''} {\Gamma'} ,\\
&\gap\gap {t'} \in \relv {k-j-1} {\Gamma''} {s''} T)\\
&\}
\end{align*}

@@

\begin{proofsketch}
For the monotonicty with regards to the step-index, this follows
directly from the definitions of $\mathcal{E}$ and $\mathcal{V}$. For the
environment and the store, this follows from the definition of $\rels
k {\Gamma \envplus{x : S}} s$. To extend the environment and the store
for $x : S$, we can append as much as we want to $\Gamma$ and $s$, to
get $\Gamma'$ and $s'$, and then ignore the last element which is for
$x : S$.
\end{proofsketch}

@@

Originally, subtyping was not regular wrt $\wfe$. Roughly,
all the $\wfe$ preconditions in subtyping were dropped. In this
broader calculus, subtyping transivitiy doesn't hold, because of the
rule (\textsc{$\sub$-rfn}) which requires expansion of the left type.

@@

Note that $u$ is realizable, since each lower bound is a subtype of
its upper bound. So it is straightforward to turn this counterexample
to subtyping transivity into a counterexample to preservation:
\begin{align*}
&\mlnew u \ldots {
\ \mlapp{\abs x \Top x}{
\ \ \mlapp{\abs f {S \tfun U} f}{
\ \ \ \mlapp{\abs f {S \tfun T} f}{
\ \ \ \ \mlapp{\abs f {S \tfun S} f}{
\ \ \  \ \ \abs x S x}}}}}
\end{align*}

>>>>

This paper presents a proposal for a new type-theoretic foundation of
Scala and languages like it. The properties we are interested in
modeling are Scala's path-dependent types and abstract type members,
as well as its mixture of nominal and structural typing through the
use of refinement types. Compared to previous approaches~\cite{nuObj,FS},
we make no attempt to model inheritance or mixin composition. Indeed
we will argue that such concepts are better modeled in a different
setting.

The calculus does not precisely describe what's currently in Scala. It
is more normative than descriptive. The main point of deviation
concerns the difference between Scala's compound type formation using
{\bf with} and classical type intersection, as it is modeled in the
calculus. Scala, and the previous calculi attempting to model it
conflates the concepts of compound types (which inherit the members of
several parent types) and mixin composition (which build classes from
other classes and traits). At first glance, this offers an economy of
concepts. However, it is problematic because mixin composition and
intersection types have quite different properties. In the case of
several inherited members with the same name, mixin composition has to
pick one which overrides the others. It uses for that the concept of
linearization of a trait hierarchy. Typically, given two independent
traits $T_1$ and $T_2$ with a common method $m$, the mixin composition
\code@$T_1$ with $T_2$@ would pick the $m$ in $T_2$, whereas the member in
$T_1$ would be available via a super-call. All this makes sense from
an implementation standpoint. From a typing standpoint it is more
awkward, because it breaks commutativity and with it several
monotonicity properties.

@@

Sets of declarations form a lattice with the given meet $\wedge$ and
join $\vee$, the empty set of declarations as the top element, and the
bottom element $\seq{D_\Bot}$, Here $\seq{D_\Bot}$ is the set of
declarations that contains for every term label $l$ the declaration
$\ldecl l \Bot$, for every type label $L$ the declaration $\Ldecl L
\Top \Bot$ and for every method label $m$ the declaration $\mdecl m
\Top \Bot$.
}

@@

\subsubsection{Set of Terms}
$\rele k \Gamma s T$ defines the set of terms that appear to have type
$T$ when taking at most $k$ steps. $s$ must agree with a {\it prefix}
of $\Gamma$, so $\Gamma$ can additionally contain variables not in
$s$. This is needed for checking methods in $\mathcal{V}$ above, and
for relating open terms. If $k > 0$, $\mathcal{E}$ extends $\Gamma$
and $s$ so that they agree. It then states that if it can reduce $t$
in the extended store to an irreducible term in $j < k$ steps, then
this term must be in a corresponding $\mathcal{V}$ set.
\begin{align*}
&\rele k \Gamma s T = \{ t \ |\\
&\gap \Gamma \ts T \wfe \andl\\
&\gap k = 0 \orl (\forall j \leq k,\\
&\gap\gap \forall (\Gamma'; s') \in \rels k \Gamma s ,\\
&\gap\gap \reductionl t {s'} {t'} {s''} j \andl\\
&\gap\gap \irred {t'} {s''} \impliesl\\
&\gap\gap \forall \Gamma'' \in \relg k {s''} {\Gamma'} ,\\
&\gap\gap {t'} \in \relv {k-j-1} {\Gamma''} {s''} T)\\
&\}
\end{align*}

@@

\begin{proofsketch}
For the monotonicity with regards to the step-index, this follows
directly from the definitions of $\mathcal{E}$ and $\mathcal{V}$. For the
environment and the store, this follows from the definition of $\rels
k {\Gamma \envplus{x : S}} s$. To extend the environment and the store
for $x : S$, we can append as much as we want to $\Gamma$ and $s$, to
get $\Gamma'$ and $s'$, and then ignore the last element which is for
$x : S$.
\end{proofsketch}

@@

Originally, subtyping was not regular wrt $\wfe$. Roughly,
all the $\wfe$ preconditions in subtyping were dropped. In this
broader calculus, subtyping transitivity doesn't hold, because of the
rule (\textsc{$\sub$-rfn}) which requires expansion of the left type.

@@

Note that $u$ is realizable, since each lower bound is a subtype of
its upper bound. So it is straightforward to turn this counterexample
to subtyping transitivity into a counterexample to preservation:
\begin{align*}
&\mlnew u \ldots {
\ \mlapp{\abs x \Top x}{
\ \ \mlapp{\abs f {S \tfun U} f}{
\ \ \ \mlapp{\abs f {S \tfun T} f}{
\ \ \ \ \mlapp{\abs f {S \tfun S} f}{
\ \ \  \ \ \abs x S x}}}}}
\end{align*}
